---
title: "CV_TestPerformance"
author: "Ming, Victor"
date: "April 4, 2017"
output: github_document
---

(Repeated CV takes a long time to run for `glmnet`. Hence to ensure knitting finishes fast, we excluded CV part for comparing performance between glmnet vs. SVM in "BuildModel_AnalyzePredictors.Rmd". CV is done in this file and CV results were saved as objects in data/R objects folder.

Step 1&2 are the same as in "BuildModel_AnalyzePredictors.Rmd", only step 3 is new.)

# S0 Set up workspace
## Load packages
```{r load packages, message=FALSE}
#source("https://bioconductor.org/biocLite.R")
#biocLite('e1071')                                    # required for glmnet in caret
#biocLite('pROC')
library(pROC)
library(ggplot2)
library(limma)
library(caret)
library(dplyr)
library(parallel)
library(doParallel)
library(readxl)
```
## Parallel processing
Set up parallel processing to speed up trian()
Make sure to specify in trainControl()

```{r parallel processing, message=FALSE}
cluster <- makeCluster(detectCores() - 1) # convention to leave 1 core for OS
registerDoParallel(cluster)
```
## Load data
Read in pre-processed data:
*Make sure the pre-processed data (data.txt, which is in data.zip) is present in the ../processed_data/ directory.

```{r load data}
setwd('../')                                           # note: all of these relative file path calls work only for knitting

# load data (pre-processed training set)
train.data <- read.table('../data/Processed Data/data.txt')
str(train.data)                                   ## row names are CpG sites, column names are sample names
# transpose our data to have rows as observations, which is more convenient later on for building models
train.data <- as.data.frame(t(train.data))

# load metadata
train.design <- read.csv("../data/Processed Data/des.txt", sep="\t", header=TRUE)
str(train.design)

row.names(train.data) == train.design$Samplename               # check that the samples are in same order
```

Read in test data:
```{r read test}
# read pre-processed test data
test.data <- read.table("../../Data/Processed Data/Test data/Matrix.processed.betas.placenta.txt", row.names = 1, header = T)
test.data <- as.data.frame(t(test.data))   #transpose data

# meta data for test data
test.design <-  read_excel("../../data/Processed Data/Test data/metadata.GA_illumina_methylation.xls", 
    sheet = "Metadata", skip = 28)

# subset only columns we need and rename them
test.design <- test.design[test.design$`Sample name` %in% rownames(test.data),]
test.design <- test.design[,c(1,7,8,10)]
colnames(test.design)[1] <- "Samplename"
colnames(test.design)[3] <- "sex"
colnames(test.design)[4] <- "ga"

str(test.design)

```

# S1 Prefiltering Features

Reducing the number of features can significantly reduce computational time, which is desirable when the dataset is large. However, we must be careful not remove potentially 'interesting' features that have a high chance of being useful in building a classifier.

## 1.1 Remove NAs

We should remove any sites with NAs in them or else predictions cannot be generated for these samples if the CpG site is chosen as a predictor.
```{r remove NAs}
#remove sites with NAs
sum(is.na(test.data)) # 52000 total entries that are NA
test.rmna <- test.data[, colSums(is.na(test.data)) == 0]  # remove columns with NAs present
```
## 1.2 Reduce CpGs to match test and train
Some CpGs were removed in the Test dataset from preprocessing and QC. These need to be removed, or errors might occur when trying to predict.

```{r filter training data to contain the same CpGs as the test}
# this isn't necessary if the test data didn't have CpGs removed (as a result of QC/preprocessing)
train.data <- train.data[,colnames(train.data) %in% colnames(test.rmna)]
```

## 1.3 Prefiltering cpgs (most variable) 
The goal of this prefiltering section is to reduce computational time without compromising detecting interesting features.

```{r prefiltering based on SD}
train.sd <- apply(as.matrix(train.data), MARGIN = 2,FUN = sd) #caculate SD for each feature
hist(train.sd)                    # histogram of the s.d.'s
abline(v = mean(train.sd)) 

# filter CpG sites with low s.d: only keep those with s.d higher than the average s.d across all CpG sites
train.gsd <- subset(train.sd, train.sd > 0.10)
hist(train.gsd)

# subset training data to only highly variable features
train.data.gsd <- train.data[,colnames(train.data) %in% names(train.gsd)]
```

We reduced the # of features to `r ncol(train.data.gsd)` to reduce computation time. `train.data.gsd` is the working dataset.
# S2 Supervised classification:
We decided to try two different models for building our classifer: elastic net logistic regression (`glmnet`) and support vector machines (SVM). Both of these models have been used in the literature to build predictive models based on 450k DNA methylation data (Horvath 2013, De Carli et al 2017), indicating that they may be well-suited for our dataset.
## 2.1 logistic regression with elastic net regularization
```{r rename}
#renamed training data for standard coding
x.train <- train.data.gsd
y.train <- train.design$Ethnicity 
```

```{r subset data for faster run time, eval = FALSE, include = FALSE}
# This subsetting is for testing code out without computational delay
# since the data is very large (~450k rows), I will subset the data first to be able to play around with the code quickly.
x.train <- train.data.gsd[,1:1000] #takes the first 1000 columns (cpgs)
```

```{r Specify resampling method}
k = 5
M = 3

fitControl <- trainControl(method = "repeatedcv", 
													 number = k,                 # Number of folds
													 repeats = M,
													 ## Estimate class probabilities
													 classProbs = TRUE,
                           ## Evaluate performance using 
                           ## the following function
                           summaryFunction = twoClassSummary,
													 allowParallel = TRUE # allow parallel processing
													 )  

netGrid <- expand.grid(alpha = c(0.75),
                           lambda = c(0.077, 0.25))
```
We specify the model to be built using repeated cross validation with a fold = 5, and repeats = 3. We tune the model holding alpha constant (alpha = 0.75), keeping alpha high to favour L1 norm to achieve a small panel of biomarkers. Lambda, the magnitude of the penalty, is tested at 0.077, and 0.25.

```{r tune glmnet parameters}
set.seed(2017)                                         # training models requires the use of random #s. Setting (set.seed()) the randomness ensures reproducibility

system.time(netFit <- train(x = x.train,   # samples need to be in rows, features need to be columns
								y = y.train,                  
								method = "glmnet",                     # glmnet model
								trControl = fitControl,                # use fitControl to specify cross validation
								tuneGrid = netGrid,
								preProcess = c( "center", "scale"),    # Center and Scale the data
								metric = 'Accuracy')                        # ROC because distribution is slightly skewed
)

netFit
netFit$results
#saveRDS(netFit, './Data/Processed Data/netFitfinal.rds')
```

```{r extract features}
predictorsNet <- predictors(netFit)
length(predictorsNet)
#write.table(predictorsNet, './Data/Processed Data/predictorsGlmnet.txt')
```
Our glmnet-built model has chosen `r length(predictorsNet)` CpGs that can be used to predict ethnicity.

## 2.2 SVM with linear kernel
This section is for building the model using SVM with a linear kernel (i.e. penalty parameter C = 1). However, because computational time is long, this section is excluded when ran, since we have chosen the glmnet model to be our final model.

```{r svm linear, eval = TRUE}
svmControl <- trainControl(method="repeatedcv",   
                           number = 5,
                           repeats=3,		    
                           summaryFunction=twoClassSummary,	# Use AUC to pick the best model
                           classProbs=TRUE,
                           allowParallel = TRUE)

system.time(svmFit <- train(x=x.train,
                            y= y.train,
                            method = "svmLinear",
                            preProc = c("center","scale"),
                            metric="ROC",
                            trControl= svmControl)	)
svmFit
```

# S3 Compare Model Performance Using Cross Validation

```{r Estimate glmnet test error by nested cv}
# this is adapted from Amrit's code from lec 19
#use repeated CV to estimate test performance
set.seed(2018)

# list of lists containing fold ids
folds <- lapply(1:M, function(i) createFolds(y.train, k = k))

system.time(netTesterror <- lapply(folds, function(i){
  lapply(i, function(j){
    # tune parameters with CV
    set.seed(2019)
    fitControl <- trainControl(method = "repeatedcv", 
													 number = k,                 
													 repeats = M,
													 classProbs = TRUE,
                           summaryFunction = twoClassSummary,
													 savePredictions = TRUE,
													 allowParallel = T)
    
    
    # build elastic net classifier
    netFit <- train(x =  x.train[-j,],   
								y = y.train[-j],                  
								method = "glmnet",                     
								trControl = fitControl,
								preProcess = c( 'center', 'scale'),
								metric = 'ROC')   
    
    # Estimate probabilities of test predictions
    probTest <- predict(netFit, x.train[j,], type = 'prob')
    ethProb <- probTest[,'Asian']
    ethProb
  })

})
)
# netTesterror
#saveRDS(netTesterror,"netTesterror.rds")
```

```{r glmnet Performance}
# Computer classification performance measures
# enet
Performance <- mapply(function(x, y){
  auc <- pROC::roc(y.train[unlist(x)], unlist(y),
                   direction ='<',
                   levels = c('Caucasian', 'Asian'),
                   percent = TRUE)
  list(tpr = auc$sensitivities,
       fpr = 100 - auc$specificities,
       auc = round(auc$auc, 2))
}, x = folds, y = netTesterror)
Performance
```

```{r ROC curve}
# plot ROC curve

plot(Performance['tpr',][[1]] ~ Performance['fpr',][[1]],
     type = 'l', col = 1, xlab = '100 - sensitivity',
     ylab = 'Sensitivity', main = 'Enet')
for(i in length(folds)){
  points(Performance['tpr',][[i]] ~ Performance['fpr',][[i]],
         type = 'l', col = 2)
}
text(x = 60, y = 40, labels =
       paste0('mean AUC = ', round(mean(unlist(Performance['auc',])), 1),
              '+/-', round(sd(unlist(Performance['auc',])), 1), '%'))
```

## Support Vector Machine

We do not need repeated CV for linear kernel SVM as it does not have tuning parameters. Regular CV is conducted for SVM instead.

```{r train svm}
# Linear kernel SVM with CV
    fitControl <- trainControl(method = "cv", 
													 number = k,                 
													 classProbs = TRUE,
                           summaryFunction = twoClassSummary,
													 savePredictions = TRUE,
													 allowParallel = T)
    
```


```{r Estimate svm test error by nested cv}
# this is adapted from Amrit's code from lec 19
#use repeated CV to estimate test performance
set.seed(2018)

# list of lists containing fold ids
folds <- createFolds(y.train, k = k)

system.time(svmTesterror <- lapply(folds, function(j){
    # tune parameters with CV
    fitControl <- trainControl(method = "cv", 
													 number = k,                 
													 classProbs = TRUE,
                           summaryFunction = twoClassSummary,
													 savePredictions = TRUE, allowParallel = T)
    
    
    # build SVM classifier
    svmFit <- train(x=x.train[-j,],
                        y= y.train[-j],
                        method = "svmLinear",
                        preProc = c("center","scale"),
                        metric="ROC",
                        trControl=fitControl)
    
    # Estimate probabilities of test predictions
    probTest <- predict(svmFit, x.train[j,], type = 'prob')
    ethProb <- probTest[,'Asian']
    ethProb

})
)
#svmTesterror
```

```{r svm Performance}
# Computer classification performance measures
# enet
Performance <- mapply(function(x, y){
  auc <- pROC::roc(y.train[unlist(x)], unlist(y),
                   direction ='<',
                   levels = c('Caucasian', 'Asian'),
                   percent = TRUE)
  list(tpr = auc$sensitivities,
       fpr = 100 - auc$specificities,
       auc = round(auc$auc, 2))
}, x = folds, y = svmTesterror)
Performance
```

```{r svm ROC curve}
# plot ROC curve

plot(Performance['tpr',][[1]] ~ Performance['fpr',][[1]],
     type = 'l', col = 1, xlab = '100 - sensitivity',
     ylab = 'Sensitivity', main = 'Linear Kernel SVM')
for(i in length(folds)){
  points(Performance['tpr',][[i]] ~ Performance['fpr',][[i]],
         type = 'l', col = 2)
}
text(x = 60, y = 40, labels =
       paste0('mean AUC = ', round(mean(unlist(Performance['auc',])), 1),
              '+/-', round(sd(unlist(Performance['auc',])), 1), '%'))
```


