---
title: "Predictive Modeling Analysis"
author: "Ming Wan, Victor Yuan"
output: 
  github_document:
    toc: TRUE
---

# Step 0: Load Packages and Data
Load required packages:


```{r}
#source("https://bioconductor.org/biocLite.R")
#biocLite('e1071')                                    # required for glmnet in caret
library(ggplot2)
library(limma)
library(caret)
library(dplyr)
library(glmnet)
```

Read in pre-processed data:
*Make sure the pre-processed data (data.txt, which is in data.zip) is present in the ../processed_data/ directory.

```{r}
setwd('../')                                           # note: all of these relative file path calls work only for knitting

# load data (pre-processed training set)
train.data <- read.table('../Data/Processed Data/data.txt')
str(train.data)
## row names are CpG sites, column names are sample names

# load metadata
design <- read.csv("../Data/Processed Data/des.txt", sep="\t", header=TRUE)
str(design)

colnames(train.data) == design$Samplename               # check that the samples are in same order
```

# Step 1: Unsupervised clustering:
---------------this section is exploratory analysis, we should move this to Nivi's exploratory file----------
As Rob suggested, PCA should be the precursor to supervised classification, more like an exploration.

## PCA on training data:

```{r pca}
pc.train <- prcomp(t(scale(t(train.data), center = T, scale = T)), center = F, scale = F)

# look at the eigenvalues
plot(pc.train) # can we add some labels here (PC1, PC2, PC3,...)
```
Say something about the plot 

```{r not sure what this chunk of code tells us, eval = FALSE, include = FALSE}
pc.train$sdev
diag((pc.train$sdev)^2)
sum(diag((pc.train$sdev)^2))
diag((pc.train$sdev)^2)[1,1]/sum(diag((pc.train$sdev)^2))
```

```{r Plot PC1 PC2 PC3}
# first 2 PCS
PC123 <- data.frame(pc.train$rotation[,c("PC1","PC2", "PC3")])              # Take out first 3 PCs
PC123 <- PC123 %>% tibble::rownames_to_column('Samplename') %>%             # Put sample names into column to 
                    left_join(design, 'Samplename')                         # Join the metadata info 
head(PC123)            

ggplot(PC123, aes(x = PC1, y = PC2)) + 
  geom_point(aes(color = Ethnicity)) +
  ggtitle('PC1 and PC2: Ethnicity')

ggplot(PC123, aes(x = PC1, y = PC3)) + 
  geom_point(aes(color = Ethnicity)) +
  ggtitle('PC1 and PC3: Ethnicity')

ggplot(PC123, aes(x = PC2, y = PC3)) + 
  geom_point(aes(color = Ethnicity)) +
  ggtitle('PC2 and PC3: Ethnicity')
```
We can see from plotting the first three principal components that our groups (Asian, Caucasian) do not seem to separate. This indicates that the main drivers of the variance in the data is something else.

```{r Plot other metadata}
ggplot(PC123, aes(x = PC1, y = PC2)) + 
  geom_point(aes(color = sex)) +
  ggtitle('Sex')

ggplot(PC123, aes(x = PC1, y = PC2)) + 
  geom_point(aes(color = ga)) +
  ggtitle('Gestational Age')

ggplot(PC123, aes(x = PC1, y = PC2)) + 
  geom_point(aes(color = Sample_Group)) +
  ggtitle('Sample Group')
```
It's not clear that our other variables are driving the variance in the data (sex, gestational age, and sample group).

```{r Scatter plot matrix first 5 PCs}
# scatter plot matrix for the first 5 PCs
splom(pc.train$rotation[,1:5], panel = panel.smoothScatter, raster = TRUE)
```
What does this plot show?

## PCA projection of loadings to test data:

```{r}
# read pre-processed test data

# project PC loadings to test data

```

--------End of exploratory Analysis-----------

# Step 2: Supervised classification:


## logistic regression with elastic net regularization

```{r Specify resampling method}
fitControl <- trainControl(method = "repeatedcv", 
													 number = 5,                 # Number of folds
													 repeats = )                 # Repeat 5 times
```

##### Compare the different methods of cross validation. Why pick repeatedcv

```{r tune glmnet}
set.seed(2017)                                         # training models requires the use of random #s. Setting (set.seed()) the randomness                                                             ensures reproducibility

# system.time measures the time it takes to run a function
system.time(netFit <- train(x = t(train.data),         # samples need to be in rows, features need to be columns
								y = design$Ethnicity, #Predicts probability of being Asian, because this is 1st level
								method = "glmnet",                     # glmnet model
								trControl = fitControl,                # use fitControl to specify repeated cross validation
								preProcess = c( "center", "scale"))    # Center and Scale the data
)

netFit
```
 
Horvath et al. (2013) uses an 'elastic net generalized linear model' to build an across-tissue DNAm predictor on age. Since our data is the same type, we'll try glmnet.

Horvath, S. (2013). DNA methylation age of human tissues and cell types. Genome Biology, 14(10), R115. http://doi.org/10.1186/gb-2013-14-10-r115

```{r logit}

```

