---
title: "Differential methylation analysis (of Processed Data) with limma"
author: "Nivretta"
output: github_document
---

#Step 0: Load Packages and Data

###Installed the necessary dependencies by: 

```{r, load libraries, message=FALSE}

#install.packages("gplots")

library(gplots)
library(limma)


```

###Load in preprocessed data from dropbox.

```{r}

#setwd("../Limma")

#download preprocessed data from this dropbox link directly into Limma folder
#data.txt is in gitignore, be sure not to try to push to github
#download.file("https://www.dropbox.com/s/s4xv0k0vsl0ka0r/data.txt?dl=1", "data.txt")


data <- read.table("data.txt")

design <- read.csv("des.txt", sep="\t", header=TRUE)

```

#Step 1: ID differentially methylated sites

We can use a linear model to identify differentially methylated probes with `limma`, as shown in this older 540 [seminar](http://www.ugrad.stat.ubc.ca/~stat540/seminars/seminar08_methylation.html).

Create a design matrix.

```{r design matrix}
des.mat <- model.matrix(~Ethnicity, design)
```

Fit model to obtain top differentially methylated CpG sites.

```{r fit linear model}

DMRfit <- lmFit(data, des.mat)
DMRfitEb <- eBayes(DMRfit)
cutoff <- 0.01
DMR <- topTable(DMRfitEb, coef = "EthnicityCaucasian", number = Inf, p.value = cutoff)
knitr::kable(head(DMR))  # top hits 

```

So using a cutoff of FDR = 0.01, we identified 106 CpG sites that are differentially methylated between Caucasian and Asian genetic ancestry. Now we can make some plots to check these hits.

#Step 3: Plotting

```{r heatmap of beta values of top 100 Limma hits}

#get the top 100 differentially methylated CpG sites & values for all samples
DMR100 <- topTable(DMRfitEb, coef = "EthnicityCaucasian", number = 100)
DMR.CpG <- t(as.matrix(subset(data, rownames(data) %in% rownames(DMR100))))
str(DMR.CpG, max.level = 0)

limma_top100_ethinicity <- list(rownames(DMR100))

write_rds(list(rownames(DMR100)), path = "limma_top100_ethinicity")

#plot heatmap
###LABELS WRONG AT THE MOMENT...
#col <- c(rep("darkgoldenrod1", times = nrow(DMR.CpG)))
#col[grepl("Caucasian", rownames(DMR.CpG))] <- "forestgreen"
#op <- par(mai = rep(0.5, 4))
#heatmap.2(DMR.CpG, col = redblue(256), RowSideColors = col, density.info = "none", 
    #trace = "none", Rowv = TRUE, Colv = TRUE, labCol = FALSE, labRow = FALSE, 
    #dendrogram = "row", margins = c(1, 5))
#legend("topright", c("Caucasian", "Asian"), col = c("darkgoldenrod1", "forestgreen"), 
    #pch = 15)
```


