---
title: "Exploratory Analysis of Processed Data"
author: "Nivretta"
output: github_document
---

#Step 0: Load Packages and Data

###Installed the necessary dependencies by: 

```{r, load libraries, message=FALSE}

#install.packages("gplots")
#install.packages("devtools")
#install_github("ggbiplot", "vqv")
library(devtools)
library(ggbiplot)
library(dplyr)
library(tibble)
library(gplots)
library(RColorBrewer)
library(lattice)

```

###Load in preprocessed data from dropbox.

```{r}
#none of the three google drive links work
#download.file("https://drive.google.com/open?id=0B3XxI9J9w-RYN0R5Rnh6SnZNNmc", "data.txt")
#download.file("https://drive.google.com/uc?export=download&id=0B3XxI9J9w-RYN0R5Rnh6SnZNNmc", "data.txt")
#download.file("https://drive.google.com/uc?export=download&confirm=kWmX&id=0B3XxI9J9w-RYN0R5Rnh6SnZNNmc", "data.txt")

#download preprocessed data from this dropbox link directly into Exploratory Analysis folder
#data.txt is in gitignore, be sure not to try to push to github
#download.file("https://www.dropbox.com/s/s4xv0k0vsl0ka0r/data.txt?dl=1", "data.txt")

#setwd("../ExploratoryAnalysis")

data <- read.table("data.txt")

design <- read.csv("des.txt", sep="\t", header=TRUE)

```

#Step 1: Exploratory Plots

A first look at the processed data with some plots.

First we need to convolve the design matrix with the processed data.

```{r Create Exploratory Analysis Data Frame}

#rename processed data columns to sample names - doesn't need to be done anymore
#colnames(gsetFin2) <- c(as.character(design$Samplename))

#join the processed data with experiemtnal design info
#note, couldn't do this on my local computer - data too large. Ran code on lab's rstudio which runs on a server
full <- cbind(design, t(data))
```

###1.1 Explore a random CpG Site


```{r Random CpG Site Plots}
#random cpg site from 1 to 464923
probe_row <- 2000

#get the site name 
probe_name <- colnames(full)[probe_row]

#plot y = beta values for random cpg site for all samples, x= gestational age, plots divided by sample group, points colored by ethnicity
ggplot(full, aes(x = as.factor(ga), y = full[probe_row], colour = Ethnicity)) + 
  #geom_boxplot(aes(fill=Ethnicity), show.legend = TRUE) + 
  geom_jitter(width = 0.5) + 
  facet_wrap(~Sample_Group) + 
  xlab("Gestational Age") + 
  ylab("Beta values") +
  ggtitle(paste("Beta values for CpG site", probe_name)) #+
  #stat_summary(fun.y = mean, geom="point", colour="darkred", size= 3)

#beta values for random cpG site, box plots of ethnicity
ggplot(full, aes(x = Ethnicity, y = full[probe_row])) + 
  geom_boxplot(aes(fill=Ethnicity), show.legend = FALSE) + 
  geom_jitter(width = 0.3) + 
  facet_wrap(~sex) + 
  xlab("Ethnicity") + 
  ylab("Beta values") +
  ggtitle(paste("Beta values for CpG site", probe_name)) +
  stat_summary(fun.y = mean, geom="point", colour="darkred", size= 3)


```

###1.2 Sample to Sample Correlations

```{r Sample to Sample Correlations}

#obtain sample names in order of ethnicity and then sample group
full_ethnicity <- full[order(full$Ethnicity, full$Sample_Group),]
order_ethnicity <- row.names(full_ethnicity)

#order expression data by ethnicity and then sample group
gsetFin2_ethnicity <- data[, order_ethnicity]

#set the color pallette for heatmap
cols<-c(rev(brewer.pal(9,"YlOrRd")), "#FFFFFF")

heatmap.2(cor(gsetFin2_ethnicity), 
          Rowv=NA, 
          Colv=NA, 
          dendrogram = "none",
          trace="none",
          col=cols, 
          margins = c(8,8),
          #labRow = c(full_time$labels),
          #labCol = c(full_time$labels),
          key.title = NA)
title("Sample Correlations, 
      Ordered by Ethnicity then sample group")

```


```{r Sample to Sample Correlations2}

#obtain sample names in order of ethnicity and then gender
full_ethnicity_sex <- full[order(full$Ethnicity, full$sex),]
order_ethnicity_sex <- row.names(full_ethnicity_sex)

#order expression data by ethnicity and then gender
gsetFin2_ethnicity_sex <- data[, order_ethnicity_sex]

#set the color pallette for heatmap
cols<-c(rev(brewer.pal(9,"YlOrRd")), "#FFFFFF")

heatmap.2(cor(gsetFin2_ethnicity_sex), 
          Rowv=NA, 
          Colv=NA, 
          dendrogram = "none",
          trace="none",
          col=cols, 
          margins = c(8,8),
          #labRow = c(full_time$labels),
          #labCol = c(full_time$labels),
          key.title = NA)
title("Sample Correlations, 
      Ordered by Ethnicity then Sex")

```


#Step 2: Unsupervised clustering

###2.1: PCA on training data:

```{r pca}
pc.train <- prcomp(scale(t(data),center=T,scale = T))

# look at the eigenvalues
plot(pc.train)
```

The `plot()` function returns a plot of the variances (y-axis) associated with the PCs (x-axis), which is useful to decide how many PCs to retain for further analysis.

```{r pca.summary}
summary(pc.train)
```

The `summary()` function describes the importance of the PCs. The first row describe again the standard deviation associated with each PC. The second row shows the proportion of the variance in the data explained by each component while the third row describe the cumulative proportion of explained variance.

```{r plot PCs}
#Plot seems unncessary and vector labels not working, ok to remove?
#g <- ggbiplot(pc.train, obs.scale = 1, var.scale = 1, 
              #groups = design$Ethnicity, ellipse = TRUE, 
              #circle = TRUE)
#g <- g + scale_color_discrete(name = '')
#g <- g + theme(legend.direction = 'horizontal', 
               #legend.position = 'top')
#print(g)



PC123 <- data.frame(pc.train$x[,1:5])              # Take out first 3 PCs
PC123 <- PC123 %>% tibble::rownames_to_column('Samplename') %>%             # Put sample names into column to 
                    left_join(design, 'Samplename')                         # Join the metadata info 
head(PC123)            

ggplot(PC123, aes(x = PC1, y = PC2)) + 
  geom_point(aes(color = Ethnicity)) +
  ggtitle('PC1 and PC2: Ethnicity')

ggplot(PC123, aes(x = PC1, y = PC3)) + 
  geom_point(aes(color = Ethnicity)) +
  ggtitle('PC1 and PC3: Ethnicity')

ggplot(PC123, aes(x = PC2, y = PC3)) + 
  geom_point(aes(color = Ethnicity)) +
  ggtitle('PC2 and PC3: Ethnicity')
```

We can see from plotting the first three principal components that our groups (Asian, Caucasian) do not seem to separate. This indicates that the main drivers of the variance in the data is something else.

```{r Plot other metadata}
ggplot(PC123, aes(x = PC1, y = PC2)) + 
  geom_point(aes(color = sex)) +
  ggtitle('Sex')

ggplot(PC123, aes(x = PC1, y = PC2)) + 
  geom_point(aes(color = ga)) +
  ggtitle('Gestational Age')

ggplot(PC123, aes(x = PC1, y = PC2)) + 
  geom_point(aes(color = Sample_Group)) +
  ggtitle('Sample Group')
```

It's not clear that our other variables are driving the variance in the data (sex, gestational age, and sample group).

```{r Scatter plot matrix first 5 PCs}
# scatter plot matrix for the first 5 PCs

splom(PC123[,c(2:6,10)], panel = panel.smoothScatter, raster = TRUE)
```

Plotting scatter plots of the top 5 PCs against ethnicity, none of the PCs can clearly separate samples by ethnicity, disappointing.

###PCA projection of loadings to test data:

We can use the predict function if we observe new data and want to predict their PCs values.

```{r}
# predict PCs for test set by projecting PC loadings to test data
predict(pc.train, new.data = test.data)
```
